An Approach for the Reduction of Boundary Effects in Time-Frequency Representations

### 2020-05-28 Notes/results.tex
paper

The experimental results show that the bias is neither depending on the noise variance $\sigma^2$ nor the forecasting length $\ell$. Indeed, independently of $sigma$, we always have $\mu_{xp}[N-1+\ell]\in\left[-0.03\sigma,0.03\sigma\right]$, which is negligible with respect to the magnitude of $\bz$. This result confirms the theoretical result~\eqref{eq:mean.error}.

On Figure~\ref{fig:res.noise.sine}, we display the experimental variance $\gamma_{xp}[N-1+\ell,N-1+\ell]$ for each value of $\sigma$ ($+$ markers). The associated theoretical asymptotic forecasting variance~\eqref{eq:cov.error.2} is also displayed in solid line. As expected, this result highlights the fact that the forecasting variance increases linearly with respect to $\sigma^2$. More surprisingly, this result shows that the forecasting variance slightly decreases with $\ell$, what is counterintuitive. It should be noted that, contrary to what expression~\eqref{eq:cov.error.2} suggests, smaller values of $\sigma$ do not  cause a decrease of the experimental variance, but an increase. This comes from the fact that when $\sigma$ is small, the matrix $\bX\bX^T$ becomes ill-conditioned. The calculation of the forecasting matrix $\tilde\bA$ in~\eqref{eq:lse} is then strongly disturbed.
\begin{figure}
%\includegraphics[width=.24\textwidth]{biasNoiseSine.eps}
\includegraphics[width=.48\textwidth]{VarianceNoiseSine.eps}
\caption{Evolution of the experimental and theoretical forecasting variance in function of the forecasting sample index for different values of $\sigma$.}
\label{fig:res.noise.sine}
\end{figure}

@@

As in the previous study, the experimental bias vanishes when $K$ increases, what confirms the approximation result~\eqref{eq:mean.error}. Besides, the experimental variance  is displayed on Figure~\ref{fig:res.size.sine} ($+$ markers), and compared with the associated theoretical variance~\eqref{eq:cov.error.2}. Each color corresponds to these experimental results obtained for a given value of $K$. This result validates the asymptotic behavior provided by~\eqref{eq:cov.error.2}, and we can show that, as soon as $K$ is sufficiently high, the product $K\gamma[\ell,\ell]$ is approximately independent of $K$.
\begin{figure}
%\includegraphics[width=.24\textwidth]{BiasKSine.eps}
\includegraphics[width=.48\textwidth]{VarianceKSine.eps}
\caption{Evolution of the experimental and theoretical forecasting variance in function of the forecasting sample index for different values of $K$.}
\label{fig:res.size.sine}
\end{figure}

@@

To highlight the fact that the linear dynamical model is sufficient to catch most of the dynamical behavior of signals following the AHM, we compare the performance of the Algorithm~\ref{alg:boundary} with reference forecasting algorithm that could be used for extending such signals. These methods are:
\begin{itemize}
\item The \emph{Extended Dynamic Mode Decomposition (EDMD)} has been developped by Williams \etal~\cite{Williams15data}. The proposed algorithm is a way to obtain an approximation of the so-called Koopman operator of the observed system, which theoretically allows to catch dynamic of nonlinear systems~\cite{Korda18linear}.
\item The \emph{Gaussian Process Regression (GPR)}~\cite{Rasmussen06gaussian} is a method relying on a probabilistic dynamical model. That one is based on the Gaussian process structure, and therefore offer more flexibility in the type of dynamic that could be modeled than the linear model~\eqref{eq:dyn.model}.
\item The \emph{Trigonometric , Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS)} method~\cite{Rasmussen06gaussian} is based on a classical decomposition of times series into a trend, a seasonal and an ARMA components, with a specific dynamic for the seasonal component. This model demands the estimation of numerous parameters and, by implication, may be slow. 
\end{itemize}

@@

\begin{table}
\centering
\caption{AHM signal. Performance of the extension methods.}
\begin{tabular}{|c||c|c|c|}
  \hline
   \multirow{2}{*}{Algorithm} & \multicolumn{2}{c|}{MSE}  & \multirow{2}{41pt}{Computing time (sec.)} \\
   \cline{2-3} & Mean & Standard deviation & \\
   \hhline{|=#=|=|=|}
   {\sf SigExt} & $1.433\times 10^{-3}$ & $4.361\times 10^{-4}$ & $0.15$ \\
   \hline
   EDMD & $3.076\times 10^{-2}$ & $8.095\times 10^{-2}$ & $2.53$\\
   \hline
   GPR & $1.436\times 10^{-3}$ & $4.346\times 10^{-4}$ & $146.33$ \\
   \hline
   TBATS & ${\color{red}X}$ & ${\color{red}X}$ & ${\color{red}X}$ \\
   \hline
\end{tabular}
\label{tab:mse.sine}
\end{table}

\subsection{Evaluation of the quality of the boundary effect reduction}

\subsubsection{Metrics}
The quality of the boundary effect reduction must be evaluated directly on the time-frequency representation. To that aim, we compare the obtained representation to the optimal representation $\ccF_N^\mathrm{opt}(\bx)$. The optimal representation is defined as the restriction of the representation of the ground-truth extended signal $\bx^\mathrm{ext}$. Therefore, we have:
\begin{equation*}
\ccF_N^\mathrm{opt}(\bx) = \cR\left( \ccF_{N+2L}(\bx^\mathrm{ext}) \right) \ .
\end{equation*}

@@

\subsubsection{Respiratory signal}
We first consider a 50 minutes-long respiratory signal. This signal is sampled at $\fs=100$~Hz. A zoom on a small portion of the signal is displayed in Figure~\ref{fig:tho}.

@@

From that large signal, we build a dataset of $48$ non-overlapping signals of 60 seconds, \ie~$N=6000$. On each of these pieces of signal, we implement the forecasting method introduced in section~\ref{ssse:res.ahm}, including the {\sf SigExt} method detailed in Algorithm~\ref{alg:extension}. However, the TBATS extension method in not implemented here because of its excessive computing time. The extensions of $7$ seconds-long on each boundary, corresponding to $L =700$. Thus, in order to catch slowly varying dynamical behaviors, the size of the training signal $M$ is chosen so that $M=\lfloor 1.5L\rfloor$. As a result of section~\ref{ssse:res.sine}, we take: $K=\lfloor2.5M\rfloor$. The average and the median of the MSE~\eqref{eq:mse} with respect to the simulations are given in Table~\ref{tab:mse}. The averaged MSE of {\sf SigExt} method is overwhelmingly higher than the averaged MSE of the other methods. This result is caused by the presence of outliers in some extensions by {\sf SigExt}. Indeed, the value median of the MSE is of the same order than those of the other methods. These outliers are probably due to the presence of pulse in the respiratory signal. That ones make the adaptive harmonic model temporary irrelevant, and break the validity of the linear dynamical model~\eqref{eq:dyn.model} used to extend the signal. That is why {\sf SigExt} is less robust to the vanishing of oscillations that the GPR or EDMD extensions.

@@

Then, we apply the boundary effects reduction method {\sf BoundEffRed} (Algorithm~\ref{alg:boundary}) on these extensions for diverse time-frequency representations: STFT, SST, RS, as well as ConceFT, an SST-based representation introduced in~\cite{Daubechies16conceft}. In Table~\ref{tab:otd.tho}, the results are compared in terms of OTD. They are also compared with the strategy consisting in a zero-padding extension of the signal. Even though the performance of the forecasting Algorithm~\ref{alg:extension} is somehow moderate, the boundary effects can be reduced dramatically on the time-frequency representations. Notice that the extension length $L$ has been set accordingly to the window length used by the time-frequency analysis tool. For instance, here the window length we use to evaluate the STFT is of 1500 samples. To prevent the STFT from being sensitive to the boundary effect, we set $L=750$. In this way, when evaluating the spectral content of the signal near its boundaries, the analysis is not limited by a lack of information all along the window support. From now on, all results are given for $L$ at equal to the half of the width of the window used in the time-frequency transform.

\begin{table}
\centering
\caption{Respiratory signal. Averaged OTD of the boundary effects reduction methods on diverse representations.}
\begin{tabular}{|>{\centering}m{35pt}||c|c|c|c|}
  \hline
   \multirow{2}{35pt}{\centering Extension method} & \multicolumn{4}{c|}{Time-Frequency Representation} \\
   \cline{2-5}
      & STFT & SST & RS & ConceFT \\
   \hhline{|=#=|=|=|=|}
   Without extension & $2.16\times 10^{-2}$ & $5.26\times 10^{-3}$ & $3.07\times 10^{-2}$ & $1.41\times 10^{-2}$ \\
   \hline
   {\sf SigExt} & $1.72\times 10^{-2}$ & $4.00\times 10^{-3}$ & $2.43\times 10^{-2}$ & $1.12\times 10^{-2}$ \\
   \hline
   EDMD & $1.75\times 10^{-2}$ & $4.23\times 10^{-3}$ & $2.45\times 10^{-2}$ & {\color{red}X} \\
   \hline
   GPR & $1.84\times 10^{-2}$ & $4.10\times 10^{-3}$ & $2.48\times 10^{-2}$ & {\color{red}X} \\
   \hline
\end{tabular}
\label{tab:otd.tho}
\end{table}

\subsubsection{Photoplethysmogram}
\label{ssse:ppg}
We perform a study similar to the previous one on a 640 second-long photoplethysmogram (PPG) signal, sampled at $\fs=125$~Hz. A 32 second-long piece of this signal is displayed on the top of Figure~\ref{fig:ppg}. The estimated 2-second extension obtained by {\sf SigExt} on both boundary of this signal is superimposed to the ground-truth signal in the bottom of Figure~\ref{fig:ppg}.

@@

We divide the signal into 32-second long pieces, and apply Algorithm~\ref{alg:boundary} on each piece. We provide in Table~\ref{tab:otd.ppg} the OTD to the optimal time-frequency representation averaged over the signals. For all the considered time-frequency representations, the results clearly shows that our algorithm reduce the influence of the boundary effect. This highlights the ability of our approach to limit the distortion due the boundary effect and provide a more accurate representations. Even though the {\sc SigExt} extension yields time-frequency  representations slightly more sensitive to boundary effects that the extensions given by EDMD or GPR, it is the only technique that allows a real-time implementation.

@@

To evaluate the influence of the noise level on the performance of {\sf BoundEffRed}, we artificially add a Gaussian noise to the measured PPG signal, whose variance is denoted $\sigma_{\mathrm{add}}^2$. It is thus an additional noise to the measurement noise actually contained in the signal. Figure~\ref{fig:otd.noise} shows the averaged OTD of {\sf BoundEffRed} for different values of $\sigma_{\mathrm{add}}^2$. At first glance, we notice that STFT is slightly more sensitive to noise than SST. On one hand, the robustness to noise of the SST is the direct consequence of the approximation result of~\cite{Daubechies16conceft}, discussed in section~\ref{sse:perf.BoundEffRed}. On the other hand, the sentivity to noise of the STFT may be interpreted as the inability to forecast the values of the time-frequency coefficients where only noise is active, while sharp time-frequency representations, such as SST, ideally vanish at these points.
\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{OTDBoundEffRed.eps}
\caption{PPG signal. Averaged OTD of {\sf BoundEffRed} in function of the noise variance.}
\label{fig:otd.noise}
\end{figure}

>>>>

The experimental results show that the bias is neither depending on the noise variance $\sigma^2$ nor the forecasting length $\ell$. Indeed, independently of $\sigma$, we always have $\mu_{xp}[N-1+\ell]\in\left[-0.03\sigma,0.03\sigma\right]$, which is negligible with respect to the magnitude of $\bz$. This result confirms the theoretical result~\eqref{eq:mean.error}.

On Fig.~\ref{fig:res.noise.sine}, we display the experimental variance $\gamma_{xp}[N-1+\ell,N-1+\ell]$ for each value of $\sigma$ ($+$ markers). The associated theoretical asymptotic forecasting variance~\eqref{eq:cov.error.2} is also displayed in solid line. As expected, this result highlights the fact that the forecasting variance increases linearly with respect to $\sigma^2$. More surprisingly, this result shows that the forecasting variance slightly decreases with $\ell$, what is counterintuitive. It should be noted that, contrary to what expression~\eqref{eq:cov.error.2} suggests, smaller values of $\sigma$ do not  cause a decrease of the experimental variance, but an increase. This comes from the fact that when $\sigma$ is small, the matrix $\bX\bX^T$ becomes ill-conditioned. The calculation of the forecasting matrix $\tilde\bA$ in~\eqref{eq:lse} is then strongly disturbed.
\begin{figure}
%\includegraphics[width=.24\textwidth]{biasNoiseSine.eps}
\includegraphics[width=.48\textwidth]{VarianceNoiseSine.eps}
\caption{Evolution of the experimental and theoretical forecasting variance in function of the forecasting sample index for different values of $\sigma$.}
\label{fig:res.noise.sine}
\end{figure}

@@

As in the previous study, the experimental bias vanishes when $K$ increases, what confirms the approximation result~\eqref{eq:mean.error}. Besides, the experimental variance  is displayed on Fig.~\ref{fig:res.size.sine} ($+$ markers), and compared with the associated theoretical variance~\eqref{eq:cov.error.2}. Each color corresponds to these experimental results obtained for a given value of $K$. This result validates the asymptotic behavior provided by~\eqref{eq:cov.error.2}, and we can show that, as soon as $K$ is sufficiently high, the product $K\gamma[\ell,\ell]$ is approximately independent of $K$.
\begin{figure}
%\includegraphics[width=.24\textwidth]{BiasKSine.eps}
\includegraphics[width=.48\textwidth]{VarianceKSine.eps}
\caption{Evolution of the experimental and theoretical forecasting variance in function of the forecasting sample index for different values of $K$.}
\label{fig:res.size.sine}
\end{figure}

@@

To highlight the fact that the linear dynamical model is sufficient to catch most of the dynamical behavior of signals following the AHM, we compare the performance of the Algorithm~\ref{alg:boundary} with reference forecasting algorithm that could be used for extending such signals. These methods are:
\begin{itemize}
\item The \emph{Extended Dynamic Mode Decomposition (EDMD)} has been developped by Williams \etal~\cite{Williams15data}. The proposed algorithm is a way to obtain an approximation of the so-called Koopman operator of the observed system, which theoretically allows to catch dynamic of nonlinear systems~\cite{Korda18linear}.
\item The \emph{Gaussian Process Regression (GPR)}~\cite{Rasmussen06gaussian} is a method relying on a probabilistic dynamical model. That one is based on the Gaussian process structure, and therefore offer more flexibility in the type of dynamic that could be modeled than the linear model~\eqref{eq:dyn.model}.
\item The \emph{Trigonometric , Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS)} method~\cite{DeLivera11forecasting} is based on a classical decomposition of times series into a trend, a seasonal and an ARMA components, with a specific dynamic for the seasonal component. This model demands the estimation of numerous parameters and, by implication, may be slow. 
\end{itemize}

@@

\begin{table}
\centering
\caption{AHM signal. Performance of the extension methods.}
\begin{tabular}{|c||c|c|c|}
  \hline
   \multirow{2}{*}{Algorithm} & \multicolumn{2}{c|}{MSE}  & \multirow{2}{41pt}{Computing time (sec.)} \\
   \cline{2-3} & Mean & Standard deviation & \\
   \hhline{|=#=|=|=|}
   {\sf SigExt} & $1.433\times 10^{-3}$ & $4.361\times 10^{-4}$ & $0.15$ \\
   \hline
   EDMD & $3.076\times 10^{-2}$ & $8.095\times 10^{-2}$ & $2.53$\\
   \hline
   GPR & $1.436\times 10^{-3}$ & $4.346\times 10^{-4}$ & $146.33$ \\
   \hline
   TBATS & $1.732\times 10^{-3}$ & $4.924\times 10^{-4}$ & $1837.12$ \\
   \hline
\end{tabular}
\label{tab:mse.sine}
\end{table}

\subsection{Evaluation of the quality of the boundary effects reduction}

\subsubsection{Metrics}
The quality of the boundary effects reduction must be evaluated directly on the time-frequency representation. To that aim, we compare the obtained representation to the optimal representation $\ccF_N^\mathrm{opt}(\bx)$. The optimal representation is defined as the restriction of the representation of the ground-truth extended signal $\bx^\mathrm{ext}$. Therefore, we have:
\begin{equation*}
\ccF_N^\mathrm{opt}(\bx) = \cR\left( \ccF_{N+2L}(\bx^\mathrm{ext}) \right) \ .
\end{equation*}

@@

\subsubsection{Respiratory signal}
We first consider a 50 minutes-long respiratory signal. This signal is sampled at $\fs=100$~Hz. A zoom on a small portion of the signal is displayed in Fig.~\ref{fig:tho}.

@@

From that large signal, we build a dataset of $48$ non-overlapping signals of 60 seconds, \ie~$N=6000$. On each of these pieces of signal, we implement the forecasting method introduced in section~\ref{ssse:res.ahm}, including the {\sf SigExt} method detailed in Algorithm~\ref{alg:extension}. However, the TBATS extension method in not implemented here because of its excessive computing time. The extensions of $7$ seconds-long on each boundary, corresponding to $L =700$. Thus, in order to catch slowly varying dynamical behaviors, the size of the training signal $M$ is chosen so that $M=\lfloor 1.5L\rfloor$. As a result of section~\ref{ssse:res.sine}, we take: $K=\lfloor2.5M\rfloor$. The average and the median of the MSE~\eqref{eq:mse} with respect to the simulations are given in Table~\ref{tab:mse}. The averaged MSE of {\sf SigExt} method is overwhelmingly higher than the averaged MSE of the other methods. This result is caused by the presence of outliers in some extensions by {\sf SigExt}. Indeed, the value median of the MSE is of the same order than those of the other methods. These outliers are probably due to the presence of pulse in the respiratory signal. That ones make the AM-FM model temporarily irrelevant, and break the validity of the linear dynamical model~\eqref{eq:dyn.model} used to extend the signal. That is why {\sf SigExt} is less robust to the vanishing of oscillations that the GPR or EDMD extensions.

@@

Then, we apply the boundary effects reduction method {\sf BoundEffRed} (Algorithm~\ref{alg:boundary}) on these extensions for diverse time-frequency representations: STFT, SST, RS, as well as ConceFT, an SST-based representation introduced in~\cite{Daubechies16conceft}. In Table~\ref{tab:otd.tho}, the results are compared in terms of OTD. They are also compared with the strategy consisting in a zero-padding extension of the signal. Even though the performance of the forecasting Algorithm~\ref{alg:extension} is somehow moderate, the boundary effects can be reduced dramatically on the time-frequency representations. Notice that the extension length $L$ has been set accordingly to the window length used by the time-frequency analysis tool. For instance, here the window length we use to evaluate the STFT is of 1500 samples. To prevent the STFT from being sensitive to the boundary effects, we set $L=750$. In this way, when evaluating the spectral content of the signal near its boundaries, the analysis is not limited by a lack of information all along the window support. From now on, all results are given for $L$ at equal to the half of the width of the window used in the time-frequency transform.

\begin{table}
\centering
\caption{Respiratory signal. Averaged OTD of the boundary effects reduction methods on diverse representations.}
\begin{tabular}{|>{\centering}m{35pt}||c|c|c|c|}
  \hline
   \multirow{2}{35pt}{\centering Extension method} & \multicolumn{4}{c|}{Time-Frequency Representation} \\
   \cline{2-5}
      & STFT & SST & RS & ConceFT \\
   \hhline{|=#=|=|=|=|}
   Without extension & $2.16\!\times\! 10^{-2}$ & $5.26\!\times\! 10^{-3}$ & $3.07\!\times\! 10^{-2}$ & $1.41\!\times\! 10^{-2}$ \\
   \hline
   {\sf SigExt} & $1.72\!\times\! 10^{-2}$ & $4.00\!\times\! 10^{-3}$ & $2.43\!\times\! 10^{-2}$ & $1.12\!\times\! 10^{-2}$ \\
   \hline
   EDMD & $1.75\!\times\! 10^{-2}$ & $4.23\!\times\! 10^{-3}$ & $2.45\!\times\! 10^{-2}$ & $1.14\!\times\! 10^{-2}$ \\
   \hline
   GPR & $1.84\!\times\! 10^{-2}$ & $4.10\!\times\! 10^{-3}$ & $2.48\!\times\! 10^{-2}$ & $1.10\!\times\! 10^{-2}$ \\
   \hline
\end{tabular}
\label{tab:otd.tho}
\end{table}

\subsubsection{Photoplethysmogram}
\label{ssse:ppg}
We perform a study similar to the previous one on a 640 second-long photoplethysmogram (PPG) signal, sampled at $\fs=125$~Hz. A 32 second-long piece of this signal is displayed on the top of Fig.~\ref{fig:ppg}. The estimated 2-second extension obtained by {\sf SigExt} on both boundary of this signal is superimposed to the ground-truth signal in the bottom of Fig.~\ref{fig:ppg}.

@@

We divide the signal into 32-second long pieces, and apply Algorithm~\ref{alg:boundary} on each piece. We provide in Table~\ref{tab:otd.ppg} the OTD to the optimal time-frequency representation averaged over the signals. For all the considered time-frequency representations, the results clearly shows that our algorithm reduce the influence of the boundary effects. This highlights the ability of our approach to limit the distortion due the boundary effects and provide a more accurate representations. Even though the {\sc SigExt} extension yields time-frequency  representations slightly more sensitive to boundary effects that the extensions given by EDMD or GPR, it is the only technique that allows a real-time implementation.

@@

To evaluate the influence of the noise level on the performance of {\sf BoundEffRed}, we artificially add a Gaussian noise to the measured PPG signal, whose variance is denoted $\sigma_{\mathrm{add}}^2$. It is thus an additional noise to the measurement noise actually contained in the signal. Fig.~\ref{fig:otd.noise} shows the averaged OTD of {\sf BoundEffRed} for different values of $\sigma_{\mathrm{add}}^2$. At first glance, we notice that STFT is slightly more sensitive to noise than SST or RS. On one hand, the robustness to noise of the SST and the RS is the direct consequence of the approximation result of~\cite{Daubechies16conceft}, discussed in section~\ref{sse:perf.BoundEffRed}. On the other hand, the sentivity to noise of the STFT may be interpreted as the inability to forecast the values of the time-frequency coefficients where only noise is active, while sharp time-frequency representations, such as SST or RS, ideally vanish at these points.
\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{OTDBoundEffRed.eps}
\caption{PPG signal. Averaged OTD of {\sf BoundEffRed} in function of the noise variance.}
\label{fig:otd.noise}
\end{figure}
